---
title: Chromatin landscape in zebrafish lymphatic system development <br> 
subtitle: ATAC-seq and RNAseq<br>data set intersection, visualisations<br>  NBIS 5931
author: Agata Smialowska
format:
  html:
    theme: cosmo
    mainfont: Georgia
    fontcolor: "#404040"
    highlight-style: pygments
    html-math-method: katex
    code-fold: true
    number-sections: true
    toc: true
    toc-title: Contents
    toc-location: left
    toc-expand: true
    toc-depth: 4
    toc-float:
      collapsed: no
    fig-width: 8
    fig-height: 8
    css: "style.css"
    code-block-bg: "grey80"
    code-block-border-left: true
    embed-resources: true
engine: knitr
date: today
bibliography: references.bib
---


Analyses for manuscript: *Panara et al, 2026*.


<br />
<br />

```{r}
#| echo: FALSE
htmltools::img(src = knitr::image_uri(file.path("logos", "nbislogo-blue-txt.png")), 
               alt = 'NBISlogo', 
               style = 'position:absolute; top:0; left:0; padding:10px; width: 10vw')
htmltools::img(src = knitr::image_uri(file.path("logos", "SLLNew.png")), 
               alt = 'SLLlogo', 
              style = 'position:absolute; top:0; right:0; padding:10px; height: 5vw')
```


```{r}
#| echo: FALSE
#| include: FALSE
#| cache: FALSE

library(knitr)


knitr::opts_chunk$set(echo = TRUE, 
                      collapse = TRUE, 
                      warning = FALSE, 
                      message = FALSE,
                      include=TRUE,
                      out.width = "80%")

options(knitr.duplicate.label = "allow")

script_pth="./src/ATACseq-RNAseq-intersection.R"
knitr::read_chunk(script_pth)
```

<br />
<br />
<br />

Sample naming convention when referring to replicates:

* pp - prox1 positive, i.e. LEC 

* no prefix - BEC


<br />
<br />


# Plots


* list of (DE) genes associated with DA chromatin (list only) (Fig 3B) **DONE**

* ATACseq signal in peaks annotated to top 50 DEGs - with RNAseq level annot (heatmap) (Fig EV3 b) **DONE**

* ATACseq DA volcano plots with LEC and BEC genes coloured (Fig 3A) **DONE**







<br />
<br />

# Environment



```{r}
#| label: prep-env
```



```{r}
#| label: dirs
```



```{r}
#| label: functions
```

<br />
<br />

## Cutoffs used in this report

```{r}
#| label: cutoffs
```

<br />
<br />

RNA-seq:

* log2FC `r rnaseq_log2FC_CO`
* fraction cells expressing `r rnaseq_pct_expr_CO`
* p adjusted `r rnaseq_padj_CO`


<br />
<br />

ATAC-seq:


* log2FC `r atac_lFC_CO` **This is different to the analysis in `ATACseq-DA` notebook where no size effect cutoff was used**

* FDR `r atac_FDR_CO`

 

<br />
<br />

# Annotations



```{r}
#| label: annots
```

<br />
<br />

# Data


<br />

RNA-seq:

```{r}
#| label: data-rnaseq
```

<br />


ATAC-seq:


```{r}
#| label:  data-atacseq
```

<br />
<br />

## Summary of ATAC-seq data

*(For reference)*

Number of peaks in DA analysis: `r nrow(atacseq_res)`



<br />
<br />

# Results


<br />
<br />


## ATAC-seq DA


```{r}
#| label:  data-atacseq
```

<br />

**This is different to the DA peaks presented in notebook ATACseq-DA**

<br />

*(For reference)*

Number of DA peaks under cutoffs used in this report:

* DA open in LEC: `r nrow(atacseq_res.lec)`

* DA open in BEC: `r nrow(atacseq_res.bec)`



<br />
<br />

### Visualisation of DARs with RNAseq based labeling

<br />
<br />


#### Figure 3A


```{r}
#| label: atacseq-volcano
```




**Figure 3A**

<br />

::: {#fig-atac-volcano}

```{r}
#| echo: false
plot_grid(volcano5_lec,volcano5_bec)
```
Volcano plot of ATAC-seq peaks coloured by the DE status of their closest feature in scRNAseq at 5dpf.
:::

<br />



#### Figure EV3 B



```{r}
#| label: atacseq-hm-DEGs
```




**Figure EV3 B**

<br />

::: {#fig-atac-hm-DEGs}

```{r}
#| echo: false
hm.da_degs
```

Heatmap of ATAC signal in peaks annotated to promoters of DEGs at 5 dpf.
:::

<br />



<br />
<br />

## Data set intersection


### Data set summary

Number of peaks, and genes annotated as their closest feature, split by peak categories is summarised in table below.



```{r}
#| label: atac-peak-gene-summary
```

<br />
<br />


**Number of genes annotated to peaks differs from the original analysis** probably due to annotation changes. The original results were:

* n=28032 peaks were significant DA, corresponding to 12839 genes.

* n=12016 peaks were preferentially open in LEC, corresponding to 5388 genes.

* n=16016 peaks were preferentially open in BEC, corresponding to 9814 genes.


<br />
<br />

<br />

::: {#tbl-atac-numbers}

```{r}
#| echo: false

kbl(peaks_stats)|>kable_styling(bootstrap_options = c("striped", "hover"))
```
Number of peaks, and genes annotated as their closest feature.
:::

<br />
<br />
<br />

```{r}
#| label: rnaseq-gene-summary
```


Summary of DE genes in RNA-seq data:

* n=`r nrow(DE_rnaseq_d5.sig)` genes were DE at indicated cutoffs;

* n=`r nrow(rnaseq_fc.d5.lec)` genes were upregulated in LEC;

* n=`r nrow(rnaseq_fc.d5.bec)` genes were upregulated in BEC; **this value was incorrect in the previous version**

<br />
<br />


### Intersection of ATAC-seq and RNA-seq



```{r}
#| label: atac-rna-intersection
```

<br />

::: {#tbl-atac-rnaseq-numbers}

```{r}
#| echo: false

kbl(table_gene_counts)|>kable_styling(bootstrap_options = c("striped", "hover"),full_width = FALSE)
```
Gene numbers up/down regulated and DA in LEC and BEC.
:::

<br />
<br />
<br />


Gene fractions up/down regulated and DA in LEC and BEC are summarised in the following table.


<br />

::: {#tbl-atac-rnaseq-numbers}

```{r}
#| echo: false

kbl(table_gene_fractions)|>kable_styling(bootstrap_options = c("striped", "hover"),full_width = FALSE)
```
Gene fractions up/down regulated and DA in LEC and BEC.
:::

<br />
<br />
<br />


### Lists for heatmaps of DE genes associated with DA chromatin

This section pertains to heatmaps “DE genes associated with DA chromatin” (Figure 3 A).

Selection:

* peaks DA in LEC annotated to LEC DE gene;

* all peaks whose DA log2FC > 1;

* applying additional log2CPM cutoff of 2, to select strongest ATAC signal;

* top 50 genes by DA logFC (largest log2FC for genes with multiple DA peaks)


<br />
<br />


#### List for Figure 3B

<br />


```{r}
#| label: list-DEGs-topDARs
```


Selected genes:

`r genes.peaks_LEC.rnaseq_LEC.top50[order(genes.peaks_LEC.rnaseq_LEC.top50)]`




<br />
<br />

# Session Info


```{r}
#| label: session-info
sessionInfo()
```

