---
title: Chromatin landscape in zebrafish lymphatic system development <br> 
subtitle: ATAC-seq<br>peak annotation, differential accessibility, visualisations and TF footprinting<br>  NBIS 5931
author: Agata Smialowska
format:
  html:
    theme: cosmo
    mainfont: Georgia
    fontcolor: "#404040"
    highlight-style: pygments
    html-math-method: katex
    code-fold: true
    number-sections: true
    toc: true
    toc-title: Contents
    toc-location: left
    toc-expand: true
    toc-depth: 4
    toc-float:
      collapsed: no
    fig-width: 8
    fig-height: 8
    css: "style.css"
    code-block-bg: "grey80"
    code-block-border-left: true
    embed-resources: true
engine: knitr
date: today
bibliography: references.bib
---


Analyses for manuscript: *Panara et al, 2026*.


<br />
<br />

```{r}
#| echo: FALSE
htmltools::img(src = knitr::image_uri(file.path("logos", "nbislogo-blue-txt.png")), 
               alt = 'NBISlogo', 
               style = 'position:absolute; top:0; left:0; padding:10px; width: 10vw')
htmltools::img(src = knitr::image_uri(file.path("logos", "SLLNew.png")), 
               alt = 'SLLlogo', 
              style = 'position:absolute; top:0; right:0; padding:10px; height: 5vw')
```


```{r}
#| echo: FALSE
#| include: FALSE
#| cache: FALSE

library(knitr)


knitr::opts_chunk$set(echo = TRUE, 
                      collapse = TRUE, 
                      warning = FALSE, 
                      message = FALSE,
                      include=TRUE,
                      out.width = "80%")

options(knitr.duplicate.label = "allow")

script_pth="./src/5931-atacseq-da.R"
knitr::read_chunk(script_pth)
```

<br />
<br />
<br />

Sample naming convention when referring to replicates:

* pp - prox1 positive, i.e. LEC 

* no prefix - BEC


<br />
<br />

# Plots

* Peak annot chipseeker plot (Fig EV 3b) **DONE**

* Venn accessible peaks (Fig EV 3C) **DONE**

* heatmap top 500 peaks by signal (Fig EV 3D) **DONE**

* heatmap top 100 DA promoters (Fig EV 3E) **DONE**

* volcano of TOBIAS LEC/BEC coloured (Fig 7A) **DONE**

* Mafb motif at promoters (Fig A4A) && lists of common / LEC / BEC promoters **DONE**

* Mafb endothelial motif at promoters (Fig A4C) && lists of common / LEC / BEC promoters **DONE**


<br />
<br />



# Environment



```{r}
#| label: prep-env
```



```{r}
#| label: dirs
```



```{r}
#| label: functions
```


```{r}
#| label: cutoffs
```

# Annotations



```{r}
#| label: annots
```




# Data



```{r}
#| label: data
```


# Results


## Peak annotation

Peaks were annotated to their closest genomic features.



```{r}
#| label: peaks_profile
```

<br />

::: {#fig-peak-TSS}

```{r}
#| echo: false
#| fig-height: 4
#| fig-width: 4

TSS_profile
```

ATAC-seq signal at TSS.
:::

<br />



```{r}
#| label: peak_annotation
```

<br />

### Figure EV 3B


**Figure EV 3B**

<br />

::: {#fig-peak-annot}

```{r}
#| echo: false

peakAnnoplot
```

Summary of ATAC-seq peak annotation.
:::

<br />


## Data exploration

Counts were scaled using TMM [@TMM] and adjusted for sorting batch using function `ComBat` from R package `sva` [@sva].

If not stated otherwise, scaled and batch adjusted logcounts were used for plotting.
PCA is presented for reference.


```{r}
#| label: TMM-norm
```


```{r}
#| label: pca
```


<br />

::: {#fig-pca}

```{r}
#| echo: false
#| fig-height: 4

pca
```
PCA plot. Logcounts after TMM scaling and sorting batch correction (ComBat) were used for PCA calculation.
:::

<br />
<br />





## Differential accessibility

Differential accessibility was inferred using `glmLRT` as implemented in `edgeR` [@edgeR].

<br />

```{r}
#| label: da
```

Summary statistics of DA peaks are presented below. Cutoffs used:

* FDR: `r FDR_CO`

* log2FC `r lFC_CO`


<br />

Peak numbers, no filtering:

* All peaks, not filtered: `r nrow(cnt_table)`

<br />


Peak numbers, after filtering off low signal peaks (i.e. used for DA analysis):


* All peaks in DA: `r nrow(DA_res)`

* Significant DA peaks at FDR  `r FDR_CO`:

* Peaks DA open in LEC: `r nrow(DA.res.sig.up)`

* Peaks DA open in BEC: `r nrow(DA.res.sig.dn)`




<br />
<br />

### Accessible peaks summary


```{r}
#| label: peaks_venn
```

#### Fig EV 3C



**Fig EV 3C** panel A




<br />

::: {#fig-venn-DA}

```{r}
#| echo: false
cowplot::plot_grid(venn_DA_peaks.0,venn_DA_peaks,
       ncol=2, labels=c("A","B"))
```

Venn diagramm of open chromatin peaks. DA peaks are at log2FC `r lFC_CO` and FDR `r FDR_CO`. A - no log2FC cutoff; B - log2FC cutoff `r lFC_CO`.
:::

<br />



## Signal Visualisation



### Top signal peaks



```{r}
#| label: peaks_highest_count
```


Top `r n_top_peaks` peaks by highest signal (average counts after TMM scaling)




#### Figure EV 3D



**Figure EV 3D**

<br />

::: {#fig-top-peaks-hm}

```{r}
#| echo: false
hm2
```

Heatmap of ATAC-seq signal in top `r n_top_peaks` peaks, by average signal. Scaled row logcounts are plotted, after adjustment for sorting batch (ComBat).
:::

<br />



### Promoter peaks


```{r}
#| label: peaks_promoter
```


```{r}
#| label: top100_promoters
```

<br />

#### Figure EV 3E


**Figure EV 3E**

<br />

::: {#fig-top-prom-hm}

```{r}
#| echo: false
hm1
```

Heatmap of ATAC-seq signal in top 100 promoters. Scaled row logcounts are plotted, after adjustment for sorting batch (ComBat).
:::

<br />




## TF footprinting

TF footprinting was performed using TOBIAS [@Bentsen2020].

<br />
<br />

### Summary volcano plot

The colouring on this plot is as in TOBIAS default: top / bottom 5 percentile of log2(score fold change) was coloured. 



```{r}
#| label: bindetect_volcano
```

<br />
<br />

#### Fig 7A


**Fig 7A**


<br />

::: {#fig-TFftprint-volcano}

```{r}
#| echo: false

volcano_diff
```

Comparison of TF footprints in LEC vs. BEC.
:::

<br />


### Mafb motif TF footprinting summary

**Please note this analysis was performed in SYMBOL gene id space.** This means that genes annotated with the SYMBOL id are included in these comparisons. 


```{r}
#| label: mafba_targets
```

#### Fig A4A

Mafb motif (consensus motifs) at promoters.


**Fig A4A**


<br />

::: {#fig-TFftprint-mafb-venn-allmotifs}

```{r}
#| echo: false

venn_mafba_combo
```

Mafb motif occupancy at promoters. Union of consensus motifs (Jaspar & Hocomoco) shown.
:::

<br />



#### Fig A4C

Mafb motif (endothelial motif) at promoters

**Fig A4C**


<br />

::: {#fig-TFftprint-mafb-venn-endothelial-motifs}

```{r}
#| echo: false

venn_mafba_endothelial
```

Mafb motif occupancy at promoters. Endothelial motif shown.
:::

<br />



<br />
<br />

# Session Info


```{r}
#| label: session-info
sessionInfo()
```
