<!-- 
This is a child document for differential consensus TAD report nbis proj 5931
version 5vi2026
based on version from 20-26i2026

HiC_ATACseq_RNAseq.qmd
HiC-ATACseq-RNAseq.R


performs feature overlap of differential TADs and boundaries
summarises the results
overlaps by permutation tests
-->



```{r include=FALSE,cache=FALSE}

library(knitr)
library(bookdown)


knitr::opts_chunk$set(echo = TRUE, 
                      collapse = TRUE, 
                      warning = FALSE, 
                      message = FALSE,
                      error=TRUE,
                      include=TRUE)

options(knitr.duplicate.label = "allow")

require(tidyverse)
```

#### Genes within selected TADs


##### LEC, BEC and vascular markers within `r txt_string_seltads`


```{r}
#| label: selTADs-gene-overlaps

subresdir_gene_pref=paste(subresdir_name,"/genes")

subresdir_gene=file.path(subresdir,"genes")
dir.create(subresdir_gene,recursive=TRUE)

lec_genes.dif_tads=get_seltad_overlaps(sel_tads.gr=sel_tads.gr,goi_genes_df=genes_lec)
bec_genes.dif_tads=get_seltad_overlaps(sel_tads.gr=sel_tads.gr,goi_genes_df=genes_bec)
vasc_genes.dif_tads=get_seltad_overlaps(sel_tads.gr=sel_tads.gr,goi_genes_df=genes_vasc)
all_genes.dif_tads=get_seltad_overlaps(sel_tads.gr=sel_tads.gr,goi_genes_df=genes_all)

save_overlap_genes(df=lec_genes.dif_tads[[2]],dir=subresdir_gene,pref1=pref_seltads,pref2="LEC_genes",cons_tads=cons_tads.dif.combined.wg)
save_overlap_genes(df=bec_genes.dif_tads[[2]],dir=subresdir_gene,pref1=pref_seltads,pref2="BEC_genes",cons_tads=cons_tads.dif.combined.wg)
save_overlap_genes(df=vasc_genes.dif_tads[[2]],dir=subresdir_gene,pref1=pref_seltads,pref2="vascular_genes",cons_tads=cons_tads.dif.combined.wg)
save_overlap_genes(df=all_genes.dif_tads[[2]],dir=subresdir_gene,pref1=pref_seltads,pref2="all_genes",cons_tads=cons_tads.dif.combined.wg)
```

<br />
<br />




**Number of genes in overlaps of `r txt_string_seltads` with genes of interest**


* LEC genes: `r nrow(lec_genes.dif_tads[[1]])`

* BEC genes: `r nrow(bec_genes.dif_tads[[1]])`

* vascular genes: `r nrow(vasc_genes.dif_tads[[1]])`

* all genes: `r nrow(all_genes.dif_tads[[1]])`


<br />
<br />


These gene lists are saved in files in directory `r subresdir_gene_pref`.

<br />
<br />


::: {#tbl-selTADs-geneannot-lec}

```{r}
#| label: tbl-selTADs-gene-overlaps-lec
#| echo: false

kbl(lec_genes.dif_tads[[1]], row.names = FALSE)|>
kable_styling(bootstrap_options = c("striped", "hover"),full_width = FALSE)

```
LEC genes within boundaries of `r txt_string_seltads`.  
:::


<br />
<br />



::: {#tbl-selTADs-geneannot-bec}

```{r}
#| label: tbl-selTADs-gene-overlaps-bec
#| echo: false

kbl(bec_genes.dif_tads[[1]], row.names = FALSE)|>
kable_styling(bootstrap_options = c("striped", "hover"),full_width = FALSE)

```
BEC genes within boundaries of `r txt_string_seltads`.  
:::



<br />
<br />

::: {#tbl-selTADs-geneannot-vasc}

```{r}
#| label: tbl-selTADs-gene-overlaps-vasc
#| echo: false

kbl(vasc_genes.dif_tads[[1]], row.names = FALSE)|>
kable_styling(bootstrap_options = c("striped", "hover"),full_width = FALSE)

```
Vascular genes within boundaries of `r txt_string_seltads`.  
:::


<br />
<br />

<br />
<br />

##### Overlap significance: `r txt_string_seltads` and DE genes



```{r}
#| label: genes-diftad-overlaps

feature_df=genes_all|>
      dplyr::rename(chr=chromosome_name,start=start_position,end=end_position,id=external_gene_name)|>
      dplyr::select(chr,start,end,id)

feature.gr=regioneR::toGRanges(as.data.frame(feature_df))


gene_annot_difTAD=feature_body_overlap(test_tads.gr=sel_tads.gr,feature.gr=feature.gr)
gene_annot_consTAD=feature_body_overlap(test_tads.gr=all_consTADs.gr,feature.gr=feature.gr)


### overlap of LEC / BEC genes with dif TADs

overlaps_difTADs.gene_lec=get_overlaps_sig(genes.all=genes_all$external_gene_name,
  genes.sel=genes_lec$external_gene_name,
  overlaps.dif=gene_annot_difTAD[["genes"]],
  overlaps.all=gene_annot_consTAD[["genes"]])

overlaps_difTADs.gene_bec=get_overlaps_sig(genes.all=genes_all$external_gene_name,
  genes.sel=genes_bec$external_gene_name,
  overlaps.dif=gene_annot_difTAD[["genes"]],
  overlaps.all=gene_annot_consTAD[["genes"]])

genes_de=c(genes_lec$external_gene_name,genes_bec$external_gene_name)

overlaps_difTADs.gene_de=get_overlaps_sig(genes.all=genes_all$external_gene_name,
  genes.sel=genes_de,
  overlaps.dif=gene_annot_difTAD[["genes"]],
  overlaps.all=gene_annot_consTAD[["genes"]])



contingency_tab_gene_lec=overlaps_difTADs.gene_lec[["cont_tab"]]
p.fisher_gene_lec=overlaps_difTADs.gene_lec[["p.fisher"]]

contingency_tab_gene_bec=overlaps_difTADs.gene_bec[["cont_tab"]]
p.fisher_gene_bec=overlaps_difTADs.gene_bec[["p.fisher"]]

contingency_tab_gene_de=overlaps_difTADs.gene_de[["cont_tab"]]
p.fisher_gene_de=overlaps_difTADs.gene_de[["p.fisher"]]
```

<br />
<br />

Statistical significance of the association of selected TADs: **`r txt_string_seltads`**
and LEC / BEC / all DE genes (gene body overlapping selected TADs)
was determined using `Fisher's exact test`.

The contingency tables are listed below.

The p value for overlap significance:

* LEC genes with gene body within selected TADs: **p=`r p.fisher_gene_lec$p.value`**.

* BEC genes with gene body within selected TADs: **p=`r p.fisher_gene_bec$p.value`**.

* all DE genes with gene body within selected TADs: **p=`r p.fisher_gene_de$p.value`**.


<br />
<br />

::: {#tbl-contingency-gene-lec}

```{r}
#| echo: false
kbl(contingency_tab_gene_lec, row.names = FALSE)|>
kable_styling(bootstrap_options = c("striped", "hover"),full_width = FALSE)
```

Contingency table for LEC genes with gene bodies within `r txt_string_seltads`.
:::

<br />

::: {#tbl-contingency-gene-bec}

```{r}
#| echo: false
kbl(contingency_tab_gene_bec, row.names = FALSE)|>
kable_styling(bootstrap_options = c("striped", "hover"),full_width = FALSE)
```

Contingency table for BEC genes with gene bodies within `r txt_string_seltads`.
:::

<br />

::: {#tbl-contingency-gene-de}

```{r}
#| echo: false
kbl(contingency_tab_gene_de, row.names = FALSE)|>
kable_styling(bootstrap_options = c("striped", "hover"),full_width = FALSE)
```

Contingency table for all DE genes (LEC and BEC) with gene bodies within `r txt_string_seltads`.
:::

<br />
<br />



#### Types of chromatin rearrangements in `r txt_string_seltads` encompassing DE and non-DE genes


Types of chromatin rearrangements in TADs encompassing DE and non-DE genes are presented on the barplot below. This plot is saved to file in the same directory as the pie chart on @fig-dif-tads-pie.


```{r}
#| label: difbnds_DEgenes-barplot

genes_selTADs=gene_annot_difTAD[[1]]
genes.LEC_selTADs=genes_selTADs|>dplyr::filter(id%in%genes_lec_dpf5.ls)
genes.BEC_selTADs=genes_selTADs|>dplyr::filter(id%in%genes_bec_dpf5.ls)

genes.DE_selTADs=genes_selTADs|>dplyr::filter(id%in%c(genes_lec_dpf5.ls,genes_bec_dpf5.ls))

selTADs.genes_all=cons_tads.dif.combined.ident
selTADs.genes_DE=cons_tads.dif.combined.ident|>dplyr::filter(tad_id%in%genes.DE_selTADs$tad_id)

nrow(selTADs.genes_all)
nrow(selTADs.genes_DE)



df_barplot1=selTADs.genes_all|>
        dplyr::select(tad_id,smpl,Type)|>
        dplyr::mutate(Gene=case_when(tad_id%in%selTADs.genes_DE$tad_id ~ "DE",
                        TRUE ~ "ns"))

df_barplot2=as.data.frame(table(df_barplot1$Gene,df_barplot1$Type))
df_barplot2=df_barplot2|>
        dplyr::rename(Gene=Var1,Type=Var2)|>
        group_by(Gene)|>
        summarise(Total=sum(Freq))

n_TADs.DE=unname(unlist(df_barplot2|>dplyr::filter(Gene=="DE")|>dplyr::select(Total)))
n_TADs.ns=unname(unlist(df_barplot2|>dplyr::filter(Gene=="ns")|>dplyr::select(Total)))


barplot_difbnds_DEgenes=ggplot(df_barplot1|>arrange(desc(Type)), aes(x=Gene,fill=Type))+
  geom_bar(position="fill", width=0.5)+
  theme(aspect.ratio = 1/0.8)+
  theme_classic()+
  theme(text = element_text(size = 16))+
  scale_fill_manual(values=c("#E69F00", "#56B4E9","#008080","#BC8F8F","#8A2BE2","#99999940"))+
  ylab("Fraction")+
  annotate("text", x=1.00, y=0.1,label=paste0("n=",n_TADs.DE), color="black",fontface = "bold")+
  annotate("text", x=2.00, y=0.1,label=paste0("n=",n_TADs.ns), color="black",fontface = "bold")

fname=paste0(file_string1,pref_seltads,".difTADbnds_barplot_DEgenes")
for (devplot in c("pdf","eps")){
        fname_full=paste(fname,devplot,sep=".")
        ggsave(plot=barplot_difbnds_DEgenes, filename=fname_full, path=pieplotdir, device=devplot)
}


barplot_difbnds_DEgenes.h=barplot_difbnds_DEgenes+
    geom_bar(position="fill", width=0.8)+
    theme(aspect.ratio = 1/4)+
    theme(legend.position="bottom")+
    coord_flip()


fname=paste0(file_string1,pref_seltads,".difTADbnds_barplot_DEgenes.horizontal")
for (devplot in c("pdf","eps")){
        fname_full=paste(fname,devplot,sep=".")
        ggsave(plot=barplot_difbnds_DEgenes.h, filename=fname_full, path=pieplotdir, device=devplot)
}

df_barplot3=as.data.frame(table(df_barplot1$Gene,df_barplot1$Type))
df_barplot3=df_barplot3|>
    dplyr::rename(Gene=Var1,Type=Var2)|>
    tidyr::pivot_wider(names_from=c(Type),values_from=Freq)|>
    dplyr::rename(StrengthChange=`Strength Change`)

df_barplot3

```

##### Figure EV2 G


<br />
<br />

**Figure EV2 G**

<br />
<br />

::: {#fig-dif-tadsbnds-DEgenes}

```{r}
#| echo: false
#| fig-height: 4

barplot_difbnds_DEgenes.h
```

Type of chromatin rearrangements on differential TAD boundaries for TADs encompassing DE and non-DE genes (ns). *n* is the total number of TADs in each group.
:::

<br />
<br />


The statistical significance of TAD boundary rearrengement type in each group (associated with DE and non-DE genes) was assesed by the `Pearson's Chi-squared test` (Chi-square goodness of fit test) and `2-sample test for equality of proportions with continuity correction` (two proportions Z test).


<br />
<br />


```{r}

tads_de=unlist(df_barplot3[df_barplot3$Gene=="DE",-1])
tads_ns=unlist(df_barplot3[df_barplot3$Gene=="ns",-1])

difTAD_DEG.chiprop=chisq.test(tads_de, tads_ns)
difTAD_DEG.chiprop$p.value

difTAD_DEG.zprop=list()

for (i in c("Complex","Merge","Shifted","Split","StrengthChange","other")){

  difTAD_DEG.zprop.i=prop.test(x=c(tads_de[names(tads_de)==i],tads_ns[names(tads_ns)==i]),n=c(n_TADs.DE,n_TADs.ns))

  difTAD_DEG.zprop[[i]]=difTAD_DEG.zprop.i$p.value

  print(paste(i,difTAD_DEG.zprop.i$p.value,sep=" "))

}

difTAD_DEG.zprop.df=do.call(rbind, difTAD_DEG.zprop)
colnames(difTAD_DEG.zprop.df)=c("p")

```

<br />
<br />

* Chi-square goodness of fit test, p value = `r difTAD_DEG.chiprop$p.value`.


* Two proportions Z test, p values:


::: {#tbl-zprop}

```{r}
#| echo: false
kbl(difTAD_DEG.zprop.df, row.names = TRUE)|>
kable_styling(bootstrap_options = c("striped", "hover"),full_width = FALSE)
```
P values from 2-sample test for equality of proportions with continuity correction (two proportions Z test) for proportions of chromatin rearrangement type at TAD boundaries associated with differential and non-differential genes.

:::

<br />
<br />



<br />
<br />


#### Overlap of selected TADs (`r txt_string_seltads`) with open chromatin

Open chromatin regions (OCRs) were queried for overlap with `r txt_string_seltads`. 
The significance was tested by (i) Fisher exact test for sets overlap and
(ii) permutation test for region association.

<br />
<br />

```{r}
#| label: atac-diftad-overlaps

#overlap of up/dn ATAC peaks with dif TADs

#DARs filtered using specificed cutoffs in `data-atac` chunk

#dfs
#atacseq_res.lec
#atacseq_res.bec

# Granges objects
#atac_peaks.up.gr
#atac_peaks.dn.gr

#all OCRs
#atacseq_res
#atac_peaks.gr


atac_peaks_up=atacseq_res.lec|>
  pull(PeakID)

atac_peaks_dn=atacseq_res.bec|>
  pull(PeakID)

atac_peaks_sig=c(atac_peaks_dn,atac_peaks_up)


overlap.atac.difTAD=feature_body_overlap(test_tads.gr=sel_tads.gr,feature.gr=atac_peaks.gr)
overlap.atac.consTAD=feature_body_overlap(test_tads.gr=all_consTADs.gr,feature.gr=atac_peaks.gr)


overlaps.difTADs.atac_up=get_overlaps_sig(genes.all=atac_da$PeakID,
  genes.sel=atac_peaks_up,
  overlaps.dif=overlap.atac.difTAD[["genes"]],
  overlaps.all=overlap.atac.consTAD[["genes"]])

contingency_tab_atac_up=overlaps.difTADs.atac_up[["cont_tab"]]
p.fisher_atac_up=overlaps.difTADs.atac_up[["p.fisher"]]

overlaps.difTADs.atac_dn=get_overlaps_sig(genes.all=atac_da$PeakID,
  genes.sel=atac_peaks_dn,
  overlaps.dif=overlap.atac.difTAD[["genes"]],
  overlaps.all=overlap.atac.consTAD[["genes"]])

contingency_tab_atac_dn=overlaps.difTADs.atac_dn[["cont_tab"]]
p.fisher_atac_dn=overlaps.difTADs.atac_dn[["p.fisher"]]


overlaps.difTADs.atac_de=get_overlaps_sig(genes.all=atac_da$PeakID,
  genes.sel=atac_peaks_sig,
  overlaps.dif=overlap.atac.difTAD[["genes"]],
  overlaps.all=overlap.atac.consTAD[["genes"]])

contingency_tab_atac_de=overlaps.difTADs.atac_de[["cont_tab"]]
p.fisher_atac_de=overlaps.difTADs.atac_de[["p.fisher"]]

```


<br />
<br />

<br />
<br />

Statistical significance of the association of selected TADs: **`r txt_string_seltads`**
and DA open chromatin regions (OCR body overlapping selected TADs)
was determined using `Fisher's exact test`.

The contingency tables are listed below.

The p value for overlap significance:

* DARs more accessible in LECs within selected TADs: **p=`r p.fisher_atac_up$p.value`**.

* DARs more accessible in BECs within selected TADs: **p=`r p.fisher_atac_dn$p.value`**.

* all DARs within selected TADs: **p=`r p.fisher_atac_de$p.value`**.


<br />
<br />

::: {#tbl-contingency-dar-lec}

```{r}
#| echo: false
kbl(contingency_tab_atac_up, row.names = FALSE)%>%
kable_styling(bootstrap_options = c("striped", "hover"),full_width = FALSE)
```

Contingency table for OCRs more accessible in LEC within `r txt_string_seltads`.
:::

<br />

::: {#tbl-contingency-dar-bec}

```{r}
#| echo: false
kbl(contingency_tab_atac_dn, row.names = FALSE)%>%
kable_styling(bootstrap_options = c("striped", "hover"),full_width = FALSE)
```

Contingency table for OCRs more accessible in BEC within `r txt_string_seltads`.
:::

<br />

::: {#tbl-contingency-dar-de}

```{r}
#| echo: false
kbl(contingency_tab_atac_de, row.names = FALSE)%>%
kable_styling(bootstrap_options = c("striped", "hover"),full_width = FALSE)
```

Contingency table for DAR OCRs within `r txt_string_seltads`.
:::

<br />

<br />
<br />


The associations between region sets (selected TADs and open chromatin peaks) 
were evaulated using `permutation tests` (n=`r n_perms` permutations), 
to address the question: are the ATAC-seq peaks (all, non-DA, LEC and BEC DA) preferentially 
associated with differential TADs. Additionally, DA chromatin regions were further grouped by their association with DEGs in LEC and BEC, to address the question whether the ATAC peaks 
preferentially accessible in LEC 
and associated with differential TADs (TADs selected for this section)
can be linked to altered gene expression.

These tests were performed jointly, i.e. `set A` (regions corresponding to TADs) was randomised only once and used for all comparisons presented in this table.


<br />

The significance of overlaps is summarised in the table below. 
Raw and corrected p values from the permutation test are listed (n=`r n_perms` permutations) in @tbl-perm-atac. 
Z score pertains to observed number of overlaps, given the distribution of number of overlaps when the query (`setA`, TADs) is randomised. Please note that z score value depends on number of regions in the query set, hence its value cannot be directly used for comparison of effect size between different tests. 
**Normalised z score** (z score divided by square root of number of regions in the query set) can be compared between different tets (@Malinverni2023). Values of -log10(adjusted p value) plotted against normalised z score are presented on @fig-perm-atac.

<br />

<br />
<br />




```{r}
#| label: atac-diftad-perm

atac_da=atacseq_res|>
  dplyr::rename(external_gene_name=SYMBOL.closest)|>
  dplyr::rename(annotation=annotation.closest)

atac_peaks.DEgex=atac_da|>
  dplyr::filter(external_gene_name%in%DE_rnaseq_d5.sig$X)

atac_peaks.LECgex=atac_da|>
  dplyr::filter(external_gene_name%in%genes_lec_dpf5.ls)

atac_peaks.BECgex=atac_da|>
  dplyr::filter(external_gene_name%in%genes_bec_dpf5.ls)



# ranges


atac_peaks.up.gr=atac_peaks.gr|>
        plyranges::filter(FDR<atac_FDR_CO,logFC>=atac_log2FC_CO)

atac_peaks.dn.gr=atac_peaks.gr|>
        plyranges::filter(FDR<atac_FDR_CO,logFC<=(-atac_log2FC_CO))

atac_peaks.sig.gr=c(atac_peaks.up.gr,atac_peaks.dn.gr)

atac_peaks.up.DEgex.gr=atac_peaks.up.gr|>
  plyranges::filter(id%in%atac_peaks.DEgex$PeakID)

atac_peaks.dn.DEgex.gr=atac_peaks.dn.gr|>
  plyranges::filter(id%in%atac_peaks.DEgex$PeakID)

atac_peaks.up.LECgex.gr=atac_peaks.up.gr|>
  plyranges::filter(id%in%atac_peaks.LECgex$PeakID)

atac_peaks.dn.BECgex.gr=atac_peaks.dn.gr|>
  plyranges::filter(id%in%atac_peaks.LECgex$PeakID)


# test

tads_peaks.lst=list()
#tads_peaks.lst[["consTADs"]]=all_consTADs.gr
tads_peaks.lst[["selTADs"]]=sel_tads.gr
tads_peaks.lst[["atac_up"]]=atac_peaks.up.gr
tads_peaks.lst[["atac_dn"]]=atac_peaks.dn.gr
tads_peaks.lst[["atac_sig"]]=atac_peaks.sig.gr
tads_peaks.lst[["atac_up_DEgex"]]=atac_peaks.up.DEgex.gr
tads_peaks.lst[["atac_dn_DEgex"]]=atac_peaks.dn.DEgex.gr
tads_peaks.lst[["atac_up_LECgex"]]=atac_peaks.up.LECgex.gr
tads_peaks.lst[["atac_dn_BECgex"]]=atac_peaks.dn.BECgex.gr

set.seed(2357)
TADs.atac.gex=crosswisePermTest(Alist=tads_peaks.lst, Blist=tads_peaks.lst, sampling=FALSE,
    gen=lengths_df,ntimes=n_perms,alternative="greater",force.parallel=TRUE, mc.set.seed=FALSE)


TADs.atac.gex_perm_df=TADs.atac.gex@multiOverlaps
atac_tads_gex.res_lst_full=bind_rows(TADs.atac.gex_perm_df, .id="setA")|>
  dplyr::select(!order.id)|>
  dplyr::rename(setB=name)|>
  dplyr::filter(!setA==setB)|>
  dplyr::select(setA,setB,p_value,adj.p_value,n_regionA,n_regionB,n_overlaps,z_score,norm_zscore,mean_perm_test,sd_perm_test)


pertest_atac_dir=file.path(subresdir,"permutation_tests_TADs_ATACseq")
dir.create(pertest_atac_dir,recursive=TRUE)


fname_pref_tads_atac=paste0(pref_seltads,"_ATACpeaks.permtest_regioneReloaded_nperm_",n_perms)

fname_perm_atac=paste0(fname_pref_tads_atac,".tsv")
save.table(df=atac_tads_gex.res_lst_full,file=fname_perm_atac,dir=pertest_atac_dir)

fname_perm_atac_rds=paste0(fname_pref_tads_atac,".rds")
fpath_rds=file.path(pertest_atac_dir,fname_perm_atac_rds)
saveRDS(TADs.atac.gex, file=fpath_rds)


atac_tads_gex.res_tbl=atac_tads_gex.res_lst_full|>
  dplyr::select(setA,setB,n_regionA,n_regionB,n_overlaps,p_value,adj.p_value,z_score,norm_zscore)|>
  dplyr::filter(!setA==setB)|>
  dplyr::filter(setA=="consTADs" | setA=="selTADs")|>
  dplyr::filter(!(setA=="consTADs" & setB=="selTADs")) |> 
  dplyr::filter(!(setB=="consTADs" & setA=="selTADs"))|>
  dplyr::mutate_if(is.numeric, format, digits=4,nsmall = 4)|>
  dplyr::rename(TADs=setA,ATACpeaks_subset=setB,n_TADs=n_regionA,n_ATAC_peaks=n_regionB)

LEC_dark="#109630"
BEC_dark="#762A83"

scale_col3=c("grey60",BEC_dark,LEC_dark)



#### 1 plot

atac_tads_gex.res_tbl.sel=atac_tads_gex.res_tbl|>dplyr::filter(TADs=="selTADs")|>
        dplyr::mutate_at(vars(p_value,adj.p_value,norm_zscore,z_score), as.numeric)|>
        dplyr::mutate(logPval=-log10(adj.p_value))|>
        dplyr::mutate(ATAC_peaks=case_when(
                str_detect(ATACpeaks_subset, regex("atac_up", ignore_case=TRUE)) ~ "LEC",
                str_detect(ATACpeaks_subset, regex("atac_dn", ignore_case=TRUE)) ~ "BEC",
                str_detect(ATACpeaks_subset, regex("atac_sig", ignore_case=TRUE)) ~ "all_DA",
                TRUE ~ NA_character_))|>
        dplyr::mutate(RNAseq=case_when(
                str_detect(ATACpeaks_subset, regex("DEgex", ignore_case=TRUE)) ~ "all_DEGs",
                str_detect(ATACpeaks_subset, regex("LECgex", ignore_case=TRUE)) ~ "LEC_DEGs",
                str_detect(ATACpeaks_subset, regex("BECgex", ignore_case=TRUE)) ~ "BEC_DEGs",
                TRUE ~ "no_RNAseq"))


TADs_atac.pl=ggplot(atac_tads_gex.res_tbl.sel,aes(x=norm_zscore,y=logPval, color=ATAC_peaks,shape=RNAseq)) +
        geom_point(size=3)+
        scale_color_manual(values=scale_col3)+
        scale_shape_manual(values = c(15,16,17,0))+
        geom_hline(yintercept=-log10(0.05),color="black",linetype=2)+
        theme_bw()+
        theme(aspect.ratio = 1/1.617)+
        xlab("Normalised Z score\n(permutation test)") +
        ylab("-log10 (p adjusted)")




fname=paste0("summary_plot.",fname_pref_tads_atac)
for (devplot in c("pdf","eps")){
        fname_full=paste(fname,devplot,sep=".")
        ggsave(plot=TADs_atac.pl, filename=fname_full, path=pertest_atac_dir, device=devplot)
}


## 2 plots

atac_tads_gex.res_tbl.sel.RNA=atac_tads_gex.res_tbl.sel|>
  dplyr::filter(!RNAseq=="no_RNAseq")

scale_col2=c(BEC_dark,LEC_dark)


TADs_atac.pl.rna=ggplot(atac_tads_gex.res_tbl.sel.RNA,aes(x=norm_zscore,y=logPval, color=ATAC_peaks,shape=RNAseq)) +
        geom_point(size=3)+
        scale_color_manual(values=scale_col2)+
        scale_shape_manual(values = c(15,16,17,0))+
        geom_hline(yintercept=-log10(0.05),color="black",linetype=2)+
        theme_bw()+
        theme(aspect.ratio = 1/1.617)+
        xlab("Normalised Z score\n(permutation test)") +
        ylab("-log10 (p adjusted)")


fname=paste0("summary_plot.GEX",fname_pref_tads_atac)
for (devplot in c("pdf","eps")){
        fname_full=paste(fname,devplot,sep=".")
        ggsave(plot=TADs_atac.pl.rna, filename=fname_full, path=pertest_atac_dir, device=devplot)
}



atac_tads_gex.res_tbl.sel.noRNA=atac_tads_gex.res_tbl.sel|>
  dplyr::filter(RNAseq=="no_RNAseq")


TADs_atac.pl.norna=ggplot(atac_tads_gex.res_tbl.sel.noRNA,aes(x=norm_zscore,y=logPval, color=ATAC_peaks)) +
        geom_point(size=3)+
        scale_color_manual(values=scale_col3)+
        scale_shape_manual(values = c(15))+
        geom_hline(yintercept=-log10(0.05),color="black",linetype=2)+
        theme_bw()+
        theme(aspect.ratio = 1/1.617)+
        xlab("Normalised Z score\n(permutation test)") +
        ylab("-log10 (p adjusted)")


fname=paste0("summary_plot.noGEX",fname_pref_tads_atac)
for (devplot in c("pdf","eps")){
        fname_full=paste(fname,devplot,sep=".")
        ggsave(plot=TADs_atac.pl.norna, filename=fname_full, path=pertest_atac_dir, device=devplot)
}


```



<br />

<br />
<br />



<br />
<br />

Complete results table was saved as file `r fname_perm_atac`. In that table, `SetA` is randomised and assessed for number of overlaps with `setB`.


<br />
<br />

<br />
<br />

::: {#tbl-perm-atac}

```{r}
#| echo: false
kbl(atac_tads_gex.res_tbl, row.names = FALSE)|>
kable_styling(bootstrap_options = c("striped", "hover"),full_width = FALSE)|>
row_spec(which(atac_tads_gex.res_tbl$adj.p_value<0.05 & atac_tads_gex.res_tbl$TADs=="selTADs"), bold = T, color = "white", background = "grey50")
```

Results of permutation tests for overlaps of reproducible and selected TADs (`r txt_string_seltads`) and open chromatin regions. ATAC-seq peaks were grouped into differentially more accessible in LEC or BEC, and additionally annotated to DEGs upregulated in their respective tissue (if prefixed by `prom` the peak was annotated to gene promoter (up to 3kb upstream of TSS)).
The alternative hypothesis was "greater". Rows with adjusted p values <0.05 are highlighted.

:::

<br />
<br />

##### Figure EV3 G

<br />
<br />

**Figure EV3 G**

<br />
<br />

::: {#fig-perm-atac}

```{r}
#| echo: false

TADs_atac.pl
```
Results of permutation tests for overlaps of reproducible and selected TADs (`r txt_string_seltads`) and open chromatin regions. ATAC-seq peaks were grouped into differentially more accessible in LEC or BEC, and additionally annotated to DEGs upregulated in their respective tissue.
The significance of these overlaps was assessed by permutation tests, where TAD regions were randomised. The horizontal line represents significance cutoff p=0.05.
:::

<br />
<br />


In addition, the plot on previous figure was split into two, based on using gene expression differences in ATAC peak selection.

<br />
<br />


::: {#fig-perm-atac2}

```{r}
#| echo: false
#| layout-ncol: 2
#| fig-cap: 
#|   - "OCR selection with gene expression taken into account"
#|   - "OCR selection without gene expression taken into account"


TADs_atac.pl.rna

TADs_atac.pl.norna

```
Results of permutation tests for overlaps of reproducible and selected TADs (`r txt_string_seltads`) and open chromatin regions. ATAC-seq peaks were grouped into differentially more accessible in LEC or BEC, and additionally annotated to DEGs upregulated in their respective tissue.
The significance of these overlaps was assessed by permutation tests, where TAD regions were randomised. The horizontal line represents significance cutoff p=0.05.
:::

<br />
<br />


<br />

<br />
<br />



<br />

<br />
<br />




